The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with \textit{Gaia} Data and Beyond

### 2019-05-10 paper/ms.tex

Before turning to numerical methods, we derive analytic expressions for the
systematic error in the actions induced from offsets in the position (e.g. the midplane or
Galactic center distance) or velocity (e.g. the local standard of rest) of the zero-point in the assumed coordinate system. We use the epicyclic approximation, which
assumes that the motion in the $z$ and $R$ components of the orbit are
decoupled and follow simple harmonic motion about a circular and planar
``guiding orbit'' \citep[][Section~3.2 and references
therein]{2008gady.book.....B}. We refer to the radius of this orbit (the
``guiding radius'') as $R_g$. This approximation is an excellent description
of thin disk-like orbits and a good description of thick disk-like orbits in
an axisymmetric potential (i.e. ignoring the influence of the Galactic bar and
spiral arms). We also make the assumption of a perfectly flat circular
velocity curve with velocity $v_c$, a good approximation near the Solar circle
\citep[e.g.][]{2017MNRAS.465...76M}.

@@

We now quantify how a systematic error in the Galactocentric coordinate system
induces phase-dependence in the actions calculated from the observed position and velocity of a star. We consider
three orbits in the model potential described in
Section~\ref{ssec:action_comp} that are typical of stars in the stellar thin
disk, stellar thick disk, and the stellar halo. We summarize their initial
positions in phase space and the actions computed by integrating their orbits
in the correct potential in Table~\ref{tab:orbits}. Each orbit, integrated
without systematic coordinate errors, is plotted in Appendix~\ref{app:orbits}.
We will refer to these orbits by their names (thin-disk, thick-disk, halo)
henceforth.

@@

Figure~\ref{fig:one_orbit_wrong_ref} shows that the actions computed in the
offset coordinate systems depend on the time (i.e. orbital phase) at which the
star's phase-space position is observed. To the observer this would
appear as time-dependence in the actions, which appear to oscillate around
their value using the correct coordinate system even though in this example
the observer is using the correctly constructed best-fit static, axisymmetric
potential. The relative size of the phase variation in each action depends on
the direction of the systematic offset as well as the true values of the
actions (i.e. the type of orbit). In reality we will have one measurement of
the phase-space position to work with, in which case the determination of the
orbital phase in $R$ or $z$ is degenerate with the degree of systematic offset
in that coordinate (see Figure~\ref{fig:cartoon}).  In the following we
therefore quote percentile ranges for the possible values computed for each
action as a proxy for the effect of these systematic errors in the coordinate
system.

@@

Gaussian summary statistics are clearly insufficient to describe the distribution shown in
Figure~\ref{fig:Jz_hist}. We therefore elect to measure this error by computing one
half the 95\uth{} percentile minus the 5\uth{} percentile of the distribution
of action values. We refer to this quantity as $\Delta J_i$ for each
action value. We plot this quantity in Figure~\ref{fig:Jz_hist} as a
horizontal arrow anchored on the true action value. Because of the bimodality
of the error distribution, this quantity roughly measures the distance from
the true action value to the peak of one of the modes. Furthermore, this
bimodality also implies that $\Delta J_i$ is not very sensitive to the exact
percentiles used. This summary statistic does not reflect the bias induced
when the midplane error is $\gtrsim z_{\text{max}}$.

@@

>>>>

Before turning to numerical methods, we derive analytic expressions for the
systematic error in the actions induced from offsets in the position (e.g. the midplane or
Galactic center distance) or velocity (e.g. the local standard of rest) of the zero-point in the assumed coordinate system. We use the epicyclic approximation, which
assumes that the motion in the $z$ and $R$ components of the orbit are
decoupled and follow simple harmonic motion about a circular and planar
``guiding orbit'' \citep[][Section~3.2 and references
therein]{2008gady.book.....B}. We refer to the radius of this orbit (the
``guiding radius'') as $R_g$. This approximation is an excellent description
of thin disk-like orbits and a good description of thick disk-like orbits in
an axisymmetric potential, i.e., ignoring the influence of the Galactic bar and
spiral arms. We also make the assumption of a perfectly flat circular
velocity curve with velocity $v_c$, a good approximation near the Solar circle
\citep[e.g.][]{2017MNRAS.465...76M}.

@@

We now quantify how a systematic error in the Galactocentric coordinate system
induces phase-dependence in the actions calculated from the observed position and velocity of a star. We consider
three orbits in the model potential described in
Section~\ref{ssec:action_comp} that are typical of stars in the thin
disk, thick disk, and halo. We summarize their initial
positions in phase space and the actions computed by integrating their orbits
in the correct potential in Table~\ref{tab:orbits}. Each orbit, integrated
without systematic coordinate errors, is plotted in Appendix~\ref{app:orbits}.
We will refer to these orbits by their names (thin-disk, thick-disk, halo)
henceforth.

@@

Figure~\ref{fig:one_orbit_wrong_ref} shows that the actions computed in the
offset coordinate systems depend on the time (i.e. orbital phase) at which the
star's phase-space position is observed. To the observer this would
appear as time-dependence in the actions, which appear to oscillate around
their value using the correct coordinate system even though in this example
appear as time-dependence in the actions, oscillating around
their value using the correct coordinate system. This time dependence comes even though
the observer is using the correctly constructed best-fit static, axisymmetric
potential. The relative size of the phase variation in each action depends on
the direction of the systematic offset as well as the true values of the
actions (i.e. the type of orbit). In reality, we will have one measurement of
the phase-space position to work with, in which case the determination of the
orbital phase in $R$ or $z$ is degenerate with the degree of systematic offset
in that coordinate (see Figure~\ref{fig:cartoon}).  In the following we
therefore quote percentile ranges for the possible values computed for each
action as a proxy for the effect of these systematic errors in the coordinate
system.

@@

Gaussian summary statistics are clearly insufficient to describe the distribution shown in
Figure~\ref{fig:Jz_hist}. We therefore elect to measure this error by computing one
half the 95\uth{} percentile minus the 5\uth{} percentile of the distribution
of action values. We refer to this quantity as $\Delta J_i$ for each
action value. We plot this quantity in Figure~\ref{fig:Jz_hist} as a
of observed action values. We plot this quantity, which we refer to as $\Delta J_i$ for each action, in Figure~\ref{fig:Jz_hist} as a
horizontal arrow anchored on the true action value. Because of the bimodality
of the error distribution, this quantity roughly measures the distance from
the true action value to the peak of one of the modes. Furthermore, this
bimodality also implies that $\Delta J_i$ is not very sensitive to the exact
percentiles used. This summary statistic does not reflect the bias induced
when the midplane error is $\gtrsim z_{\text{max}}$.

@@

\section{Coordinate System Offsets II. Measurement Errors} \label{sec:mes_err}
We consider the second potential source of coordinate system offsets which
would generate the action errors described in Section~\ref{sec:ref_frame}:
uncertainties in measurements of the position and velocity of the Sun relative
to the Galactic center. Converting from heliocentric to Galactocentric
coordinates relies on these measurements. Therefore, errors in their values
will induce a systematic offset in the Galactocentric phase-space position of
any observed star.

\subsection{Galactic Center}
First, one must define the center of the Galaxy. This is usually taken to be
the location of the central supermassive black hole, Sagittarius~A\textsuperscript{*}
\citep[\sgra{}, e.g.][]{2004ApJ...616..872R}. From stellar motions near
\sgra{}, the distance from the Sun to \sgra{}, referred to as $R_0$, can be
precisely measured \citep{2009ApJ...692.1075G, 2018AA...615L..15G}, with a
recent measurement using near-infrared interferometry of $8.178 \pm 0.035$~kpc
\citep{2019arXiv190405721A}.

In the epicyclic approximation (Section~\ref{ssec:epi_action}), the $0.3\%$
error in $R_0$ induces a $5.4\%$ and $5.9\%$ error in the thin-disk and
thick-disk orbits, respectively. For $J_{\phi}$, the error is $0.3\%$ for both
orbits.

\subsection{Galactic Orientation}
Second, one must define the angular orientation of the Galaxy. This was
defined in 1958 by the IAU subcomission 33b \citep{1960MNRAS.121..123B} by
defining the coordinates of the Galactic center in B1950 coordinates as
(17:42:26.6, -28:55:00) and the North Galactic pole as (12:49:00, +27:24:00).
These two quantities, together with $R_0$, define the orientation of the
Galactic plane. However, there is growing evidence that the stellar midplane
is tilted relative to this coordinate system \citep{2014ApJ...797...53G,
2016ARAA..54..529B}, though not the H~\textsc{ii} midplane
\citep{2019ApJ...871..145A}.

\subsection{Solar Height}
Third, one must define the Sun's vertical distance from the Galactic midplane, $z_{\odot}$.
It can be determined by identifying where the stellar density and
velocities reach a maximum (effectively the median height of all disk stars).
The Solar height is usually taken to be $\sim 25\,\pc$
\citep{2001ApJ...553..184C}, with a more recent measurement from \textit{Gaia}
DR2 placing it at $20.8 \pm 0.3\,\pc$ \citep{2019MNRAS.482.1417B}. Another
strategy is to use the cold gas or H~\textsc{ii} regions in the disk to define
the Galactic midplane, leading to slightly different values (by $\sim 5\,\pc$)
for the Sun's relative height \citep[e.g.][]{2019ApJ...871..145A}. A pre-{\em
Gaia} review of these measurements is given by \citet{2016ARAA..54..529B}.

Under the epicyclic approximation, the highest precision measurement of
$z_{\odot}$ of $1.4\%$ from \citet{2019MNRAS.482.1417B} induces a $0.5\%$ and
$0.04\%$ error in $J_z$ for the thick-disk and thin-disk orbits, respectively.
The discrepancy between the stellar and gas midplanes of $\sim5\,\pc$ induces
an $8.3\%$ and $1.2\%$ error in $J_z$ for the two orbits.

\subsection{Local Standard of Rest}
Finally, one must define the LSR, or mean velocity of stars near the Sun
relative to the Galactic center (which is defined to have zero velocity), and
the velocity of the Sun relative to the LSR. The radial ($U_{\odot}$) and
vertical ($W_{\odot}$) components are computed by taking the mean motions of
different stellar groups \citep[e.g.][]{2012MNRAS.427..274S}. The azimuthal
component ($V_{\odot}$) is more difficult to measure, but can be modeled using
the asymmetric drift relation \citep{2008gady.book.....B}. The values of the
components of the LSR are usually taken from \citet{2010MNRAS.403.1829S}. The
value of the circular velocity is taken to be $\sim 220\kms$
\citep[e.g.][]{2012ApJ...759..131B}.

Considerable effort has been placed on each of these measurements, but
uncertainties remain, and detailed dynamical modeling across the disk ---
particularly for dynamically cold stars --- may have to take them into
account. For instance, the discrepancy of $\sim 5\,\pc$ between the midplane
defined by stars and by gas will induce a $\sim10\%$ error in $J_z$ for an
orbit with $z_{\text{max}}\sim100\,\pc$ (see Section~\ref{sec:ref_frame}).
